# Chapter 5 - Data Transformation {-}

Load the libraries needed for these exercises.

```{r 5-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r 5-library, message=FALSE}
library(tidyverse)
library(nycflights13)
```


## 5.2 - Filter Rows with filter() {-}

### Problem 1 {-}

Find all flights that:

* Had an arrival delay of two or more hours

```{r 5-2-1a}
flights %>%
  filter(arr_delay >= 120) %>%
  count()
```

* Flew to Houston (IAH or HOU)

```{r 5-2-1-b}
flights %>%
  filter(dest %in% c('IAH', 'HOU')) %>%
  count()
```

* Were operated by United, American, or Delta

```{r 5-2-1c}
flights %>%
  filter(carrier %in% c('UA', 'AA', 'DL')) %>%
  count()
```

* Departed in summer (July, August, and September)

```{r 5-2-1d}
flights %>%
  filter(month %in% c(7, 8, 9)) %>%
  count()
```

* Arrived more than two hours late, but didn’t leave late

```{r d-2-1e}
flights %>%
  filter(arr_delay >= 120, dep_delay <= 0) %>%
  count()
```

* Were delayed by at least an hour, but made up over 30 minutes in flight

```{r 5-2-1f}
flights %>%
  filter(dep_delay >= 60, arr_delay <= dep_delay - 30) %>%
  count()
```

* Departed between midnight and 6am (inclusive)

```{r 5-2-1g}
flights %>%
  filter(dep_time >= 0, dep_time <= 600) %>%
  count()
```

### Problem 2 {-}

Another useful dplyr filtering helper is `between()`. What does it do? Can you 
use it to simplify the code needed to answer the previous challenges?

`between()` is a shortcut for `x >= left & x <= right`. We can simplify the last 
answer to Problem 1 as:

```{r 5-2-2}
flights %>%
  filter(between(dep_time, 0, 600)) %>%
  count()
```

### Problem 3 {-}

How many flights have a missing dep_time? What other variables are missing? 
What might these rows represent?

We use `is.na()` to filter the flights with a missing departure time.

```{r 5-2-3a}
flights %>%
  filter(is.na(dep_time)) %>%
  count()
```

Using `summary()` to see the breakout of the other variables, there appear to 
be flights that were cancelled.

```{r 5-2-3b}
flights %>%
  filter(is.na(dep_time)) %>%
  summary()
```

### Problem 4 {-}

Why is NA ^ 0 not missing? Why is NA | TRUE not missing? 
Why is FALSE & NA not missing? Can you figure out the general rule? 
(NA * 0 is a tricky counterexample!)

Working through these examples:
  * Anything to the zero power is 1
  * Anything OR `TRUE` is `TRUE`
  * Anything AND `FALSE` is `FALSE`

These results apply no matter what the LHS side, and so will apply to NA as well.

```{r 5-2-4a}
NA ^ 0
NA | TRUE
NA & FALSE
```

However operations on NA will return NA. NA * 0 is counter intuitive since 
you would think that anything multiplied by 0 would be 0.

```{r 5-2-4b}
NA * 0
NA ^ 2
NA + 1
```

## 5.3 Arrange Rows with arrange() {-}

### Problem 1 {-}

How could you use arrange() to sort all missing values to the start? 
(Hint: use is.na()).

We can sort missing values using the format:

```{r 5-3-1}
flights %>%
  arrange(desc(is.na(dep_time))) %>%
  head()
```

### Problem 2 {-}

Sort flights to find the most delayed flights. Find the flights that left 
earliest.

The most delayed flights (by `arr_delay`) are:

```{r 5-3-2a}
flights %>%
  arrange(desc(arr_delay)) %>%
  head()
```

The flights that left earliest (by `dep_delay`) are:

```{r 5-3-2b}
flights %>%
  arrange(dep_delay) %>%
  head()
```

### Problem 3 {-}

Sort flights to find the fastest flights.

We first calculate average speed in MPH as distance / hours in the air, and sort 
on the calculated variable.

```{r 5-3-3}
flights %>%
  mutate(speed = distance / (air_time / 60)) %>%
  arrange(desc(speed)) %>%
  select(speed) %>%
  head()
```

### Problem 4 {-}

Which flights traveled the longest? Which traveled the shortest?

The longest flights are:

```{r 5-3-4a}
flights %>%
  arrange(desc(distance)) %>%
  head()
```

The shortest flights are:

```{r 5-3-4b}
flights %>%
  arrange(distance) %>%
  head()
```

## 5.4 Select Columns with select() {-}

### Problem 1 {-}

Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, 
`arr_time`, and `arr_delay` from flights.

We can put all the variables directly into our `select()` statement:

```{r 5-4-1a}
flights %>%
  select(dep_time, dep_delay, arr_time, arr_delay) %>%
  head()
```

Or this would be a good place to try the `starts_with()` function: 

```{r 5-4-1b}
flights %>%
  select(starts_with("dep"), starts_with("arr"))
```

Or we can try a regex using `matches()`

```{r 5-4-1c}
flights %>%
  select(matches("^dep"), matches("^arr")) %>%
  head()
```

### Problem 2 {-}

What happens if you include the name of a variable multiple times in a 
select() call?

Duplicating a variable within `select()` will still just return that variable 
one time:

```{r 5-4-2}
flights %>%
  select(dep_time, dep_time) %>%
  head()
```

### Problem 3 {-}

What does the one_of() function do? Why might it be helpful in conjunction 
with this vector?

`r vars <- c("year", "month", "day", "dep_delay", "arr_delay")`

`one_of()` allows you select variables from within a character vector. We can 
pass `vars` to select everything from the vector:

```{r 5-4-3}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")

flights %>%
  select(one_of(vars))
```

### Problem 4 {-}

Does the result of running the following code surprise you? How do the select 
helpers deal with case by default? How can you change that default?

```{r 5-4-4a}
select(flights, contains("TIME")) %>%
  head()
```

`contains()` contains an argument `ignore.case` which defaults to `TRUE`, we can 
set this to `FALSE` if needed:

```{r 5-4-4b}
select(flights, contains("TIME", ignore.case = FALSE)) %>%
  head()
```

## 5.5 - Add New Variables with mutate() {-}

### Problem 1 {-}

Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to 
compute with because they’re not really continuous numbers. Convert them to a 
more convenient representation of number of minutes since midnight.

Use the modular arithmetic operators to break the time into its hours and 
minute components:

```{r 5-5-1}
flights %>%
  select(dep_time, sched_dep_time) %>%
  mutate(dep_time_cont = ((dep_time %/% 100) * 60 + (dep_time %% 100)),
         sched_dep_time_cont = ((sched_dep_time %/% 100) * 60 + (sched_dep_time %% 100))) %>%
  head()
```

### Problem 2 {-}

Compare `air_time` with `arr_time - dep_time`. What do you expect to see? What do 
you see? What do you need to do to fix it?

The issue is that `air_time` is in minutes, while `arr_time` and `dep_time` 
are not:

```{r 5-5-2a}
flights %>%
  mutate(air_time_derived = arr_time - dep_time) %>%
  select(air_time, air_time_derived) %>%
  head()
```

Instead use the `mutate()` statement from Problem 1, however the two differ:

```{r}
flights %>%
  mutate(dep_time_cont = ((dep_time %/% 100) * 60 + (dep_time %% 100)),
         arr_time_cont = ((arr_time %/% 100) * 60 + (arr_time %% 100)),
         air_time_derived = arr_time_cont - dep_time_cont) %>%
  select(air_time, air_time_derived) %>%
  head()
```

### Problem 3 {-}

Compare dep_time, sched_dep_time, and dep_delay. How would you expect those 
three numbers to be related?

We would expect `dep_delay` to be the difference between the `dep_time` and the 
`sched_dep_time`. But be sure to convert from time to continuous first:

```{r 5-5-3}
flights %>%
  mutate(dep_time_cont = ((dep_time %/% 100) * 60 + (dep_time %% 100)),
         sched_dep_time_cont = ((sched_dep_time %/% 100) * 60 + (sched_dep_time %% 100)),
         dep_delay_derived = dep_time_cont - sched_dep_time_cont) %>%
  select(dep_delay, dep_delay_derived) %>%
  head()
```

### Problem 4 {-}

Find the 10 most delayed flights using a ranking function. How do you want to 
handle ties? Carefully read the documentation for `min_rank()`.

We'll use `min_rank()` to rank the flights by `arr_delay`:

```{r 5-5-4}
flights %>%
  select(arr_delay) %>%
  mutate(most_delayed = min_rank(-arr_delay)) %>%
  filter(most_delayed <= 10) %>%
  arrange(most_delayed)
```

### Problem 5 {-}

What does `1:3 + 1:10` return? Why?

We get an error because `1:3` + `1:10` are not multiples of each other:

```{r 5-5-5a}
1:3 + 1:10
```

Think through what is happening under the hood. This operation is recycling the 
shorter vector:

  * 1 + 1
  * 2 + 2
  * 3 + 3
  * 4 + 1
  * 5 + 2
  * 6 + 3
  * 7 + 1
  * 8 + 2
  * 9 + 3
  * 10 + 1 - error because `1:3` has not been fully cycled through

So the following will not return an error:

```{r 5-5-5b}
1:3 + 1:12
```

### Problem 6 {-}

What trigonometric functions does R provide?

`R` has the following trig functions within `base`:

  * cos(x)
  * sin(x)
  * tan(x)

  * acos(x)
  * asin(x)
  * atan(x)
  * atan2(y, x)

  * cospi(x)
  * sinpi(x)
  * tanpi(x)
  
Note that angles are given in radians:

```{r}
cos(pi * 0.25)

# cospi(x) is equivalent to cos(pi * x)

cospi(0.25)
```

##  5.6 Grouped summaries with summarise() {-}

### Problem 1 {-}

Brainstorm at least 5 different ways to assess the typical delay characteristics 
of a group of flights. Consider the following scenarios:

* A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the 
time.

```{r 5-6-1a}
flights %>%
  group_by(flight) %>%
  summarise(fifteen_early = mean(arr_delay <= -15, na.rm = TRUE),
            fifteen_late = mean(arr_delay >= 15, na.rm = TRUE)) %>%
  filter(fifteen_early == 0.50,
         fifteen_late == 0.50)
```

* A flight is always 10 minutes late.

```{r 5-6-1b}
flights %>%
  group_by(flight) %>%
  summarise(ten_late = mean(arr_delay == 10, na.rm = TRUE)) %>%
  filter(ten_late == 1.00)
```

* A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of 
the time.

```{r 5-6-1c}
flights %>%
  group_by(flight) %>%
  summarise(thirty_early = mean(arr_delay <= -30, na.rm = TRUE),
            thirty_late = mean(arr_delay >= 30, na.rm = TRUE)) %>%
  filter(thirty_early == 0.50,
         thirty_late == 0.50)
```

* 99% of the time a flight is on time. 1% of the time it’s 2 hours late.

```{r 5-6-1d}
flights %>%
  group_by(flight) %>%
  summarise(on_time = mean(arr_delay <= 0, na.rm = TRUE),
            late = mean(arr_delay >= 120, na.rm = TRUE)) %>%
  filter(on_time == 0.99,
         late == 0.01)
```

### Problem 2 {-}

Come up with another approach that will give you the same output as 
`not_cancelled %>% count(dest)` and 
`not_cancelled %>% count(tailnum, wt = distance)` (without using count()).

First setup the `not_cancelled` data set.

```{r 5-6-2a}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))
```

The first chunk of code gives us the following output:

```{r 5-6-2b}
not_cancelled %>%
  count(dest)
```

We can replicate this without using `count()` by doing a `group_by()` on `dest`:

```{r 5-6-2c}
not_cancelled %>%
  group_by(dest) %>%
  summarise(n = n())
```

The second chunk of code gives us:

```{r 5-6-2d}
not_cancelled %>% 
  count(tailnum, wt = distance)
```

Again we can avoid using `count` by doing a `group_by()` on `tailnum`. Since 
`wt = distance` gives the total number of miles flown, we use `sum()` instead:

```{r 5-6-2e}
not_cancelled %>%
  group_by(tailnum) %>%
  summarise(n = sum(distance))
```

### Problem 3 {-}

Our definition of cancelled flights `(is.na(dep_delay) | is.na(arr_delay) )` is 
slightly suboptimal. Why? Which is the most important column?

`arr_delay` is the more important of the two columns - filtering on `arr_delay`
alone will give the same subset: 

```{r 5-6-3}
flights %>%
  summarise(suboptimal = sum(is.na(dep_delay) | is.na(arr_delay)),
            optimal = sum(is.na(arr_delay)))
```

### Problem 4 {-}

Look at the number of cancelled flights per day. Is there a pattern? Is the 
proportion of cancelled flights related to the average delay?

There is a generally positive trend between average delay and the proportion of 
cancelled flights, with a couple of outliers.

```{r 5-6-4a}
flights %>%
  mutate(date = lubridate::make_date(year, month, day)) %>%
  group_by(date) %>%
  summarise(cancelled = mean(is.na(arr_delay)),
            avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  ggplot(aes(avg_delay, cancelled)) +
  geom_point()
```

A lot of flights were cancelled on February 8th and 9th, although the average 
delays those days were not that large - a snowstorm hit the region that weekend, 
with a lot of flights preemptively cancelled.

```{r 5-6-4b}
flights %>%
  mutate(date = lubridate::make_date(year, month, day)) %>%
  group_by(date) %>%
  summarise(cancelled = mean(is.na(arr_delay)),
            avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(cancelled)) %>%
  head()
```

### Problem 5 {-}

Which carrier has the worst delays? Challenge: can you disentangle the effects 
of bad airports vs. bad carriers? Why/why not? (Hint: think about 
`flights %>% group_by(carrier, dest) %>% summarise(n()`)

At first glance, Frontier and AirTran seem to have the worst delays.

```{r 5-6-5a}
flights %>%
  group_by(carrier) %>%
  summarise(median_delay = median(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(median_delay))
```

We can try to get a better sense of bad airlines vs bad airports by grouping by
both, but this will be thrown off by `carrier`-`dest` combinations that occur 
infrequently.

```{r 5-6-5b}
flights %>%
  group_by(carrier, dest) %>%
  summarise(median_delay = median(arr_delay, na.rm = TRUE)) %>%
  ggplot(aes(carrier, median_delay)) +
  geom_boxplot()
```

### Problem 6 {-}

What does the `sort` argument to `count()` do. When might you use it?

The `sort` argument will arrange `count()` in descending order. If we quickly 
wanted to find the most popular destinations, we could do:

```{r 5-6-6}
flights %>%
  group_by(dest) %>%
  count(sort = TRUE) %>%
  head()
```

## 5.7 - Grouped Mutates (and Filters) {-}

### Problem 1 {-}

Refer back to the lists of useful mutate and filtering functions. Describe how 
each operation changes when you combine it with grouping.

### Problem 2 {-}

Which plane `(tailnum)` has the worst on-time record?

We do a familiar `group_by()`-`summarise()` to calculate the proportion of 
flights with an `arr_delay` less than or equal to 0, and then apply a `filter` 
to see the on-time performance of planes with more than twenty flights.

```{r 5-7-2}
flights %>%
  group_by(tailnum) %>%
  summarise(flights = n(),
            on_time = mean(arr_delay <= 0, na.rm = TRUE)) %>%
  select(tailnum, flights, on_time) %>%
  filter(flights > 20) %>%
  arrange(on_time) %>%
  head()
```

### Problem 3 {-}

What time of day should you fly if you want to avoid delays as much as possible?

It appears that flights later in day have a greater chance of being delayed 
than those early in the morning.

```{r 5-7-3}
flights %>%
  group_by(hour) %>%
  filter(!is.na(dep_delay)) %>%
  summarise(delayed = mean(dep_delay > 0, na.rm = TRUE)) %>%
  ggplot(aes(x = hour, y = delayed)) +
  geom_col()
```

### Problem 4 {-}

For each destination, compute the total minutes of delay. For each flight, 
compute the proportion of the total delay for its destination.

A grouped mutate comes in handy here, as we can first calculate the total 
minutes of delay for each destination and then use that value to compute the 
proportion of the total delay attributable to each flight.

```{r 5-7-4}
flights %>% 
  group_by(dest) %>%
  filter(arr_delay > 0) %>%  
  mutate(total_delay = sum(arr_delay),
         prop_delay = arr_delay / sum(arr_delay)) %>%
  select(dest, flight, total_delay, prop_delay) %>%
  arrange(desc(total_delay)) %>%
  head()
```

### Problem 5 {-}

Delays are typically temporally correlated: even once the problem that caused 
the initial delay has been resolved, later flights are delayed to allow earlier 
flights to leave. Using lag() explore how the delay of a flight is related to 
the delay of the immediately preceding flight.

First group by `origin` and the perform a `mutate()` with `lag()` to get the 
delay of the immediately preceding flight. We then use `summarise` to get the 
correlation between the delay and lagged delay for each airport.

```{r 5-7-5}
flights %>%
  group_by(origin) %>%
  filter(!is.na(dep_delay)) %>%
  arrange(year, month, day, hour, minute) %>%
  mutate(lag_delay = lag(dep_delay)) %>%
  summarise(delay_correlation = cor(dep_delay, lag_delay, use = 'complete.obs'))
```

### Problem 6 {-}

Look at each destination. Can you find flights that are suspiciously fast? 
(i.e. flights that represent a potential data entry error). Compute the air 
time a flight relative to the shortest flight to that destination. Which 
flights were most delayed in the air?

A grouped `mutate` is helpful here, as we can calculate the mean air time by 
destination and then immediately use that value to calculate a flight's 
deviation from it. Note that we filtered out flights less than an hour long.

```{r 5-7-6}
deviation <- flights %>%
  group_by(dest) %>%
  filter(!is.na(air_time)) %>%
  mutate(mean_air_time = mean(air_time),
         deviation = (air_time - mean_air_time) / mean_air_time) %>%
  filter(mean_air_time > 60) %>%
  arrange(deviation) %>%
  select(air_time, mean_air_time, deviation, origin, dest)

deviation %>% head()
deviation %>% tail()
```

### Problem 7 {-}

Find all destinations that are flown by at least two carriers. Use that 
information to rank the carriers.

We first use a `group_by` with `mutate()` and `filter()` to subset the 
destinations serviced by at least 2 carriers, followed by a second `group_by` 
to rank the carriers by total destinations served. We also merge data from 
the `airlines` data set to get the full carrier names.

ExpressJet and Endeavor Air are regional airlines which operate as American 
Eagle, United Express, and Delta Connection.

```{r 5-6-7}
flights %>%
  group_by(dest) %>%
  mutate(carriers = n_distinct(carrier)) %>%
  filter(carriers >= 2) %>%
  group_by(carrier) %>%
  summarise(destinations = n_distinct(dest)) %>%
  arrange(desc(destinations)) %>%
  left_join(airlines) %>%
  select(carrier, name, destinations) %>%
  head()
```

### Problem 8 {-}

For each plane, count the number of flights before the first delay of greater 
than 1 hour.

First `group_by()` `tailnum` and then apply a `filter()` using the `cumall` 
window function to subset everything before the first delay of more than 1 hour.

```{r 5-7-8}
flights %>%
  arrange(year, month, day, hour, minute) %>%
  group_by(tailnum) %>%
  filter(cumall(dep_delay <= 60)) %>%
  count() %>%
  arrange(desc(n)) %>%
  head()
```

