# Chapter 10 - Tibbles {-}

Load the libraries needed for these exercises.

```{r 10-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r 10-library, message=FALSE}
library(tidyverse)
```

## 10.5 - Exercises {-}

### Problem 1 {-}

How can you tell if an object is a tibble? (Hint: try printing `mtcars`, which 
is a regular data frame).

Use the `is.tibble()` function to determine if an object is a tibble. A tibble 
will also have a heading if printed to the console.

```{r 10-5-1}
is.tibble(mtcars)
is.tibble(diamonds)
diamonds
```

### Problem 2 {-}

Compare and contrast the following operations on a `data.frame` and 
equivalent tibble. What is different? Why might the default data frame 
behaviours cause you frustration?

A data frame will attempt to auto-complete, while a tibble will not:

```{r 10-5-2a}
df <- data.frame(abc = 1, xyz = "a")
tib <- tibble(abc = 1, xyz = "a")

df$x
tib$x
```

Data frames will sometimes simplify:

```{r 10-5-2b}
is.data.frame(df[, "xyz"])
is.tibble(tib[, "xyz"])
```

Tibble will display useful information on the data:

```{r 10-5-2c}
df[, c("abc", "xyz")]
tib[, c("abc", "xyz")]
```


### Problem 3 {-}

If you have the name of a variable stored in an object, e.g. `var <- "mpg"`, 
how can you extract the reference variable from a tibble?

Use `[]` to extract the variable and `[[]]` to extract the vector:

```{r 10-5-3}
var <- 'mpg'
as.tibble(mtcars)[var]
as.tibble(mtcars)[[var]]
```

### Problem 4 {-}

Practice referring to non-syntactic names in the following data frame by:

1. Extracting the variable called 1.

```{r 10-5-4a}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)

annoying %>%
  select(`1`)
```


2. Plotting a scatterplot of 1 vs 2.

```{r 10-5-4b}
ggplot(annoying, aes(`1`, `2`)) +
  geom_point()
```

3. Creating a new column called 3 which is 2 divided by 1.

```{r 10-5-4c}
annoying <- annoying %>%
  mutate(`3` = `2` / `1`)
```


4. Renaming the columns to one, two and three.

```{r 10-5-4d}
annoying %>%
  select(one = `1`,
         two = `2`,
         three = `3`)
```

### Problem 5 {-}

What does `tibble::enframe()` do? When might you use it?

`enframe()` can be used to convert a vector or list to a tibble:

```{r 10-5-5}
enframe(c(a = 5,  b = 10))
```

### Problem 6 {-}

What option controls how many additional column names are printed at the 
footer of a tibble?

The `tibble.max_extra_cols` option controls this behavior, with a default of 
100.