# Chapter 11 - Data Import {-}

Load the libraries needed for these exercises.

```{r 11-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r 11-library, message=FALSE}
library(tidyverse)
```

## 11.2 - Getting Started {-}

### Problem 1 {-}

What function would you use to read a file where fields were separated with
“|”?

Use `read_delim()`, using `|` as the delimiter:

```{r 11-2-1}
data <- 'a|b|c\n1|2|3'
read_delim(data, delim = '|')
```

### Problem 2 {-}

Apart from `file`, `skip`, and `comment`, what other arguments do `read_csv()` 
and `read_tsv()` have in common?

`read_csv()` and `read_tsv()` are essentially just `read_delim` with the 
delimiter preset to either a comma or a tab. All of their other arguments are 
the same.

### Problem 3 {-}

What are the most important arguments to `read_fwf()`?

The most important argument to `read_fwf()` is `col_positions`, as this 
determines how data is read:

```{r 11-2-3}
fwf_sample <- readr_example("fwf-sample.txt")
cat(read_lines(fwf_sample))
read_fwf(fwf_sample, fwf_widths(c(20, 10, 12), c("name", "state", "ssn")))
read_fwf(fwf_sample, fwf_widths(c(5, 10, 12), c("name", "state", "ssn")))

```

### Problem 4 {-}

Sometimes strings in a CSV file contain commas. To prevent them from causing 
problems they need to be surrounded by a quoting character, like `"` or `'`. 
By convention, `read_csv()` assumes that the quoting character will be `"`, 
and if you want to change it you’ll need to use `read_delim()` instead. 
What arguments do you need to specify to read the following text 
into a data frame?

`"x,y\n1,'a,b'"`

Since `read_delim()` must be used instead of `read_csv()`, the delimiter must 
be set. The `quote` argument can be set to a single quote instead of a 
double quote:

```{r 11-2-4}
data <- "x,y\n1,'a,b'"
read_delim(data, delim = ',', quote = '\'')
```

### Problem 5 {-}

Identify what is wrong with each of the following inline CSV files. What happens 
when you run the code?

* There are more data than columns, which results in a parsing failure. 
The extra data are dropped from the data frame:

```{r 11-2-5a}
read_csv("a,b\n1,2,3\n4,5,6")
```

* There is too little data in row 2 and too much data in row 3 - row 2 is filled 
in with a missing value while row 3 drops data:

```{r 11-2-5b}
read_csv("a,b,c\n1,2\n1,2,3,4")
```

* There are two variables but only one data point - `b` is filled in with a 
missing value:

```{r 11-2-5c}
read_csv("a,b\n\"1")
```

* Appears that the header was entered twice, so the data are parsed as character 
instead of a string. Or if the goal here was to enter a missing value `NA`, 
note that the `n` was processed as a new line `\n`. 

```{r 11-2-5d}
read_csv("a,b\n1,2\na,b")
read_csv("a,b\n1,2\nna,b")
```

* `read_csv()` has a delimiter set to `,`, use `read_csv2()` instead:

```{r 11-2-5e} 
read_csv("a;b\n1;3")
read_csv2("a;b\n1;3")
```

## 11.3 - Parsing a Vector {-}

### Problem 1 {-}

What are the most important arguments to `locale()`?

The `date_names` argument provides useful defaults for a `locale()` object:

```{r 11-3-1}
dutch <- locale('nl')
japanese <- locale('ja')

str(dutch)
str(japanese)
```

Be sure to read the full documentation for `locale()`. Common data import issues 
can probably be solved with `decimal_mark`, `grouping_mark`, and/or `encoding`.

### Problem 2 {-}

What happens if you try and set `decimal_mark` and `grouping_mark` to the same 
character? What happens to the default value of `grouping_mark` when you set 
`decimal_mark` to `“,”`? What happens to the default value of `decimal_mark` when 
you set the grouping_mark to `“.”`?

`locale()` requires that `decimal_mark` and `grouping_mark` be different:

```{r 11-3-2a, error=TRUE}
x <- locale(decimal_mark = '.', grouping_mark = '.')
```

Setting `decimal_mark` to `,` will automatically update `grouping_mark` to `.`. 
Similarly setting `grouping_mark` to `.` will automatically update `decimal_mark` 
to `,`:

```{r 11-3-2b}
x <- locale(decimal_mark = ',')
x$grouping_mark

y <- locale(grouping_mark = '.')
y$decimal_mark
```

### Problem 3 {-}

I didn’t discuss the `date_format` and `time_format` options to `locale()`. 
What do they do? Construct an example that shows when they might be useful.

A specific `date_format` and `time_format` structure can be specified in a 
`locale()`. This can be useful for formatting data with non-standard formatting:

```{r 11-3-3}
parse_date('June/7/90', locale = locale(date_format = '%B/%d/%y'))
parse_time('1:15PM', locale = locale(time_format = '%I:%M%p'))
```

### Problem 4 {-}

If you live outside the US, create a new locale object that encapsulates the 
settings for the types of file you read most commonly.

Create a locale with the time zone updated:

```{r 11-3-4} 
str(locale(tz = 'US/Eastern'))
```

### Problem 5 {-}

What’s the difference between `read_csv()` and `read_csv2()`?

`read_csv()` has a delimiter set to `,` while `read_csv2` is set to `;`, as 
some countries use `,` as the `decimal_mark`:

```{r 11-3-5}
read_csv('a,b\n1,2')
read_csv2('a;b\n1;2')
```

### Problem 6 {-}

What are the most common encodings used in Europe? What are the most common 
encodings used in Asia? Do some googling to find out.

See [this list](https://en.wikipedia.org/wiki/Character_encoding#Common_character_encodings) 
of common encodings (via Wikipedia). 

### Problem 7 {-}

Generate the correct format string to parse each of the following dates and times:

```{r 11-3-7a}
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014
t1 <- "1705"
t2 <- "11:15:10.12 PM"
```

Build up a datetime format using the pieces described in the chapter:

```{r 11-3-7b}
parse_date(d1, format = '%B %d, %Y')
parse_date(d2, format = '%Y-%b-%d')
parse_date(d3, format = '%d-%b-%Y')
parse_date(d4, format = '%B %d (%Y)')
parse_date(d5, format = '%m/%d/%y')
parse_time(t1, format = '%H%M')
parse_time(t2, format = '%I:%M:%OS %p')
```

