# Chapter 12 - Tidy Data {-}

Load the libraries needed for these exercises.

```{r 12-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r 12-library, message=FALSE}
library(tidyverse)
```

## 12.2 Tidy Data {-}

### Problem 1 {-}

Using prose, describe how the variables and observations are organised in 
each of the sample tables.

`table1` is the 'tidy' dataset: each variable has its own column, each 
observation has its own row, and each value has its own cell:

```{r 12-2-1a}
table1
```

`table2` combines `cases` and `population` into one column called `type`, this 
means that each variable does not have its own column, and that each observation 
spans multiple rows:

```{r 12-2-1b}
table2
```

In `table3` the variable `rate` violates a tidy principle, with multiple values 
contained in a cell, which also means that each variable does not have its own 
column:

```{r 12-2-1c}
table3
```

`table4a` and `table4b` separate cases and population into their own tables 
across years, with multiple observations in each row:

```{r 12-2-1d}
table4a
table4b
```

### Problem 2 {-}

Compute the rate for `table2`, and `table4a + table4b`. You will need to perform 
four operations:

- Extract the number of TB cases per country per year.
- Extract the matching population per country per year.
- Divide cases by population, and multiply by 10000.
- Store back in the appropriate place.

Which representation is easiest to work with? Which is hardest? Why?

**NOTE** these exercises demonstrate the difficulties of working with non-tidy 
data, the methods to come later in this chapter will greatly simplify the below 
code.

Create a `tidy` version of `table2` by filtering `type` into two tables and 
using the `dplyr` `full_join()` function to recreate `table1`

```{r 12-2-2a}
table2a <- table2 %>%
  filter(type == 'cases') %>%
  select(country, year, cases = count)

table2b <- table2 %>%
  filter(type == 'population') %>%
  select(country, year, population = count)

full_join(table2a, table2b) %>%
  mutate(rate = cases / population * 10000)
```

Use similar logic on `table4a` and `table4b` - this ends up being a bit more 
work as the data are already stored across two tables:

```{r 12-2-2b}
table4a_1 <- table4a %>%
  mutate(year = 1999) %>%
  select(country, year, cases = `1999`)

table4a_2 <- table4a %>%
  mutate(year = 2000) %>%
  select(country, year, cases = `2000`)

table4b_1 <- table4b %>%
  mutate(year = 1999) %>%
  select(country, year, population = `1999`)

table4b_2 <- table4b %>%
  mutate(year = 2000) %>%
  select(country, year, population = `2000`)

bind_rows(table4a_1, table4a_2) %>%
  full_join(bind_rows(table4b_1, table4b_2)) %>%
  mutate(rate = cases / population * 10000)
```

### Problem 3 {-}

Recreate the plot showing change in cases over time using `table2` instead of 
`table1`. What do you need to do first?

Filter `type` so that only `cases` is plotted:

```{r 12-2-3}
table2 %>%
  filter(type == 'cases') %>%
  ggplot(aes(year, count)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

## 12.3 - Spreading and Gathering {-}

### Problem 1 {-}

Why are `gather()` and `spread()` not perfectly symmetrical? Carefully consider 
the following example:

```{r 12-3-1a}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)

stocks

stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`)
```

(Hint: look at the variable types and think about column names.)
Both `spread()` and `gather()` have a `convert` argument. What does it do?

In the above example `gather()` and `spread()` are not perfectly symmetrical as 
`year` is converted from a numeric to a character variable. Use the `convert` 
argument to automatically run `type.convert()` on the `key` column:
 
```{r 12-3-1b}
stocks %>% 
  spread(year, return, convert = TRUE) %>% 
  gather("year", "return", `2015`:`2016`, convert = TRUE)
```

### Problem 2 {-}

Why does this code fail?

```{r 12-3-2a, error=TRUE}
table4a %>% 
  gather(1999, 2000, key = "year", value = "cases")
```

Be sure to use backticks to include a nonstandard variable name within a 
`tibble`:

```{r 12-3-2b}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```

### Problem 3 {-}

Why does spreading this `tibble` fail? How could you add a new column to fix 
the problem?

```{r 12-3-3a}
people <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
```

Spreading the given `tibble` will fail because rows 1 and 3 are identical 
observations. Add a `count` variable to fix the problem:

```{r 12-3-3b}
people$count <- c(1,1,2,1,1)

people %>%
  spread(key, value)
```

### Problem 4 {-}

Tidy the simple `tibble` below. Do you need to `spread` or `gather` it? What are 
the variables?

```{r 12-3-4a}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
```

The given `tibble` is not tidy as one variable (`sex`) is spread across 
multiple columns. Use `gather()`:

```{r 12-3-4b}
preg %>%
  gather(male:female, key = sex, value = count)
```


