# Chapter 20 - Vectors {-}

Load the libraries needed for these exercises.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE}
library(tidyverse)
```

## 20.3 - Important types of atomic vector {-}

### Problem 1 {-}

Describe the difference between is.finite(x) and !is.infinite(x).

``` {r 21.3.1}
# Given a character vector, is.finite() returns a vector of FALSEs and !is.infinite() returns a vector of FALSEs: 

x <- c("a", "b", "c")

is.finite(x)
!is.infinite(x)
```

### Problem 2 {-}

Read the source code for dplyr::near() (Hint: to see the source code, drop the ()). How does it work?

``` {r 21.3.2}
dplyr::near

# near() takes two numeric vectors, x and y, and checks to see if the absolute value of their difference is less than 1.490116e-08. 
```

### Problem 3 {-}

A logical vector can take 3 possible values. How many possible values can an integer vector take? How many possible values can a double take? Use google to do some research.

An integer vector can take two possible values: any integer number or NA. A double vector can take 5 possible values: any real number, NA, NaN, Inf, or -Inf. 

### Problem 4 {-}

Brainstorm at least four functions that allow you to convert a double to an integer. How do they differ? Be precise.

``` {r 21.3.4}
x <- c("a" = 2.1,"b" = 2.4,"c" = 4.3,"d" = 5.8)
x <- as.integer(x)
mode(x)
class(x)
#as.integer() strips attributes, such as names. It changes the class of the vector to integer, but does not change the object's mode, the way R stores the object.

x <- c("a" = 2.1,"b" = 2.4,"c" = 4.3,"d" = 5.8)
(class(x) <- "integer")
mode(x)
#This approach retains the vector's names, changes the class to integer, and leaves the mode unchanged.

x <- c("a" = 2.1,"b" = 2.4,"c" = 4.3,"d" = 5.8)
(storage.mode(x) <- "integer")
class(x)
#This approach retains names and changes the object's mode and class to integer. 

x <- c("a" = 2.1,"b" = 2.4,"c" = 4.3,"d" = 5.8)
(mode(x) <- "integer")
class(x)
#This approach retains names and changes the object's mode and class to integer. (Unsure how this differs from storage.mode())
```

### Problem 4 {-}

What functions from the readr package allow you to turn a string into logical, integer, and double vector?

The parse_* functions take a character vector and parse it into a logical, integer, or double vector. 

## 20.4 - Using atomic vectors {-}

### Problem 1 {-}

What does mean(is.na(x)) tell you about a vector x?

``` {r 20.4.1a}
#The function mean(is.na(x)) returns the proportion of NA values in the vector x: 
x <- c(1, 2, 3, NA)
mean(is.na(x))
```

What about sum(!is.finite(x))?

``` {r 20.4.1b}
#The function returns the number of non-finite values in the vector x:
x <- c(1, 2, 3, NA, Inf)
sum(!is.finite(x))
```

### Problem 2 {-}

Carefully read the documentation of is.vector(). What does it actually test for? 

``` {r 20.4.2a}
# Is vector tests whether the object is a vector with no attributes other than names. Therefore, it returns false if the object is a factor:

y <- factor(c(1, 2, 3, 4))
is.vector(y)

# It returns true if the object is a non-atomic vector, but only if the mode is "any": 

x <- list(c(1, 2, 3), c("a", "b", "c"))
is.vector(x)

x <- list(c(1, 2, 3))
is.vector(x, "numeric")
```

Why does is.atomic() not agree with the definition of atomic vectors above?


``` {r 20.4.2b}
# The above definition deems NULL objects to be of a different type than lists or atomic vectors, but is.atomic() returns TRUE if passed a NULL object:

x <- NULL
is.atomic(x)
```

### Problem 3 {-}

Compare and contrast setNames() with purrr::set_names().

``` {r 20.4.3}
# Both functions will take a vector and assign it the names included in another vector:

x <- c(1, 2, 3, 4)
y <- c("a", "b", "c", "d")

set_names(x, y)

setNames(x, y)

# But in addition to stricter argument checking, set_names() will name a vector after itself if not given a vector of names: 

set_names(x)

# And accepts naming functions: 

set_names(1:10, function(z) letters[seq_along(z)])
```

### Problem 4 {-}

Create a function that takes a vector as input and returns the last value. (Should you use [ or [[?)

``` {r 20.4.4a}

last_val <- function(x) {
  x[[length(x)]]
}

# Double brackets are better than single brackets, because they return the last value of a list as a vector rather than another list:

x <- list(c(1, 2, 3, 6), (c("a", "b", "c")))
last_val(x)
```

Create a function that takes a vector as input and returns the elements at even numbered positions.

``` {r 20.4.4b}

even_val <- function(x) {
  y <- seq(2, length(x), 2)
  x[y]
}

x <- c(1, 3, 5, 7, 9)

even_val(x)
```

Create a function that takes a vector as input and returns every element except the last value.

``` {r 20.4.4c}

first_vals <- function(x) {
  x[-length(x)]
}

x <- c(1, 3, 5, 7, 9)

first_vals(x)
```

Create a function that takes a vector as input and returns only even numbers (and no missing values).

``` {r 20.4.4d}

evens <- function(x) {
  y <- x[!is.na(x)]
  y[y %% 2 == 0]
}

x <- c(1, 3, 4, 5, NA)

evens(x)
```



