---
output:
  html_document: default
  pdf_document: default
---
# Chapter 21 - Iteration {-}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r library, message=FALSE}
library(tidyverse)
library(stringr)
library(purrr)
```

## 21.2 For loops {-}

### Problem 1 {-}

Write a for loop to compute the mean of every column in mtcars.

``` {r 21.2.1a}
x <- vector("double", ncol(mtcars))
for (i in seq_along(mtcars)) {
  x[[i]] <- mean(mtcars[[i]])
}
```

Write a for loop to determine the type of each column in nycflights13::flights.

``` {r 21.2.1b}
x <- vector("character", ncol(nycflights13::flights))
for (i in seq_along(nycflights13::flights)) {
  x[[i]] <- str_c(class(nycflights13::flights[[i]]), collapse = ", ")
}
```

Write a for loop to compute the number of unique values in each column of iris.

``` {r 21.2.1c}
x <- vector("integer", ncol(iris))
for (i in seq_along(iris)) {
  x[[i]] <- length(unique(iris[[i]]))
}
```

Write a for loop to generate 10 random normals for each of mean = -10, 0, 10, 100. 
``` {r 21.2.1d}
means <- c(rep(-10, 10), rep(0, 10), rep(10, 10), rep(100, 10))
x <- vector("double", 40)
for (i in seq_along(means)) {
  x[[i]] <- rnorm(1, means[[i]])
}
```

### Problem 2 {-}

Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:

``` {r 21.2.2a}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}

str_c(letters[1:26], collapse = "")
```

``` {r 21.2.2b}
x <- sample(100)
sd <- 0
for (i in seq_along(x)) {
  sd <- sd + (x[i] - mean(x)) ^ 2
}
sd <- sqrt(sd / (length(x) - 1))

sd(x)
```

``` {r 21.2.2c}
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}

cumsum(x)
```

### Problem 3 {-}

Write a for loop that prints() the lyrics to the children’s song “Alice the camel”.

``` {r 21.2.3a}
humps <- c("five", "four", "three", "two", "one", "no")

y <- vector("character", 6)
for (i in seq_along(humps)) {
  x <- str_replace_all(str_c(c(rep("Alice the camel has x humps; ", 3), "So go, Alice, go. "), collapse = ""), "x", humps)
}
print(str_c(c(x, " Now Alice has no humps!"), collapse = ""))
```

Convert the nursery rhyme “ten in the bed” to a function. Generalise it to any number of people in any sleeping structure.

``` {r 21.2.3b}
roll_bed <- function(n, structure) {
  
n <- n:2 
y <- vector("numeric", length(n))
for (i in seq_along(n)) {
  y <- str_c("There were ",n," in the ",structure," and the little one said, 'Roll over! Roll over!' So they all rolled over and  none fell out.")
}
y <- c(y, str_c("There was 1 in the ", structure, " and the little one said, 'Alone at last!'"))
 print(y)
}
roll_bed(10, "bed")
```

Convert the song “99 bottles of beer on the wall” to a function. Generalise to any number of any vessel containing any liquid on any surface.

``` {r 21.2.3c}
beer_wall <- function(n, liquid) {

n <- n:1
y <- vector("numeric", length(n))
for (i in seq_along(n)) {
  y <- str_replace_all(str_c(n, " bottles of beer on the wall, ", n, " bottles of beer. Take one down and pass it around, ", (n-1)," bottles of beer on the wall."), "beer", liquid)
}
n <- n[1]
y <- c(y, str_replace_all(str_c("No more bottles of beer on the wall, no more bottles of beer. Go to the store and buy some more, ", n, " bottles of beer on the wall."), "beer", liquid))
 print(y)
}
beer_wall(10, "beer")
```

### Problem 4 {-}

It’s common to see for loops that don’t preallocate the output and instead increase the length of a vector at each step:
``` {}
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output
```
How does this affect performance? Design and execute an experiment.

``` {r 21.2.4}
data <- nycflights13::flights %>% 
  select(-year, -carrier, -tailnum, -origin, -dest, -time_hour)

library(microbenchmark)

# With preallocation
x <- vector("numeric", 13)
microbenchmark(for (i in seq_along(data))
  x[[i]] <- mean(data[[i]])
)

#  Without preallocation
y <- vector("numeric", 0)
microbenchmark(for (i in seq_along(data))
  y[[i]] <- mean(data[[i]])
)
```

## 21.3 For loop variations {-}

### Problem 1 {-}

Imagine you have a directory full of CSV files that you want to read in. You have their paths in a vector, files <- dir("data/", pattern = "\\.csv$", full.names = TRUE), and now want to read each one with read_csv(). Write the for loop that will load them into a single data frame.

``` {r 21.3.1}
files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)
output <- vector("list", length(files))
for (i in seq_along(files)) {
  output[[i]] <- read_csv(files[[i]])
}
output <- purrr::flatten_df(output)
```

### Problem 2 {-}

What happens if you use for (nm in names(x)) and x has no names? 

``` {r 21.3.2a}
no_iris <- unname(iris)
no_iris <- no_iris[, 1:4]

# Leaves the list empty and does not throw an error.

y <- vector("list")
for (nm in names(no_iris)) {
  y[[nm]] <- mean(no_iris[[nm]])
}
y
```

What if only some of the elements are named? 

``` {r 21.3.2b}

# Returns a list of length 2 with the value of the named element. 

names(no_iris[[1]]) <- "Sepal.length"
y <- vector("list")
for (nm in names(no_iris)) {
  y[[nm]] <- mean(no_iris[[nm]])
}
y
```

What if the names are not unique?

``` {r 21.3.2c}
# Only returns the mean of the first element and does not throw an error.

names(no_iris) <- c("Sepal.length", "Sepal.length", "Sepal.length", "Sepal.length")
y <- vector("list")
for (nm in names(no_iris)) {
  y[[nm]] <- mean(no_iris[[nm]])
}
y
```

### Problem 3 {-}

Write a function that prints the mean of each numeric column in a data frame, along with its name. For example, show_mean(iris) would print:

show_mean(iris)
#> Sepal.Length: 5.84
#> Sepal.Width:  3.06
#> Petal.Length: 3.76
#> Petal.Width:  1.20
(Extra challenge: what function did I use to make sure that the numbers lined up nicely, even though the variable names had different lengths?)

``` {r 21.3.3}
print_means <- function(data) {
data <- select_if(data, is.numeric)

Means <- vector("double", length(data))
for (i in seq_along(data)) {
  Means[[i]] <- mean(data[[i]], na.rm = TRUE)
}

Names <- names(data)

print(as_data_frame(cbind(Names, Means)))
}
```

### Problem 4 {-}

What does this code do? How does it work?

``` {r 21.3.4}
trans <- list( 
  disp = function(x) x * 0.0163871,
  am = function(x) {
    factor(x, labels = c("auto", "manual"))
  }
)
for (var in names(trans)) {
  mtcars[[var]] <- trans[[var]](mtcars[[var]])
}

# The list trans contains two functions: disp and am. Disp takes its input and  multiplies by .0163871 and am takes its input and creates a factor variable with the value "auto" or "manual." The for loop loops over the two functions in the list, uses the variables disp" and am in the dataset mtcars as inputs, and outputs a new mtcars dataset.
```

## 21.4 For loops vs. functionals {-}

### Problem 1 {-}

Read the documentation for apply(). In the 2d case, what two for loops does it generalise?

``` {r 21.4.1}
# Unsure. 
```

### Problem 2 {-}

Adapt col_summary() so that it only applies to numeric columns You might want to start with an is_numeric() function that returns a logical vector that has a TRUE corresponding to each numeric column.

``` {r 21.4.2}
# Original col_summary: 
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}

# Adapted col_summary:
col_summary <- function(df, fun) {
  df <- select_if(df, is.numeric)
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}
```

## 21.5 The map functions

### Problem 1

Write code that uses one of the map functions to compute the mean of every column in mtcars.

``` {r 21.5.1a}
mtcars %>% 
  map_dbl(mean)
```

Write code that uses one of the map functions to determine the type of each column in nycflights13::flights

``` {r 21.5.1b}
nycflights13::flights %>% 
  map_chr(~str_c(class(.), collapse = ", "))
```

Write code that uses one of the map functions to compute the number of unique values in each column of iris

``` {r 21.5.1c}
iris %>% 
  map_int(~n_distinct(.))
```

Write code that uses one of the map functions to generate 10 random normals for each of μ = -10, 0, 10, and 100. 

``` {r 21.5.1d}
x <- c(-10, 0, 10, 100)
x %>% 
  map(~rnorm(10, mean = .))
```

### Problem 2

How can you create a single vector that for each column in a data frame indicates whether or not it’s a factor?

``` {r 21.5.2}
factor_fun <- function(data) {
  data %>%
    map_lgl(~is.factor(.))
}
```

### Problem 3

What happens when you use the map functions on vectors that aren’t lists? What does map(1:5, runif) do? Why?

``` {r 21.5.3}
#The function iterates across each value, 1:5, in the vector and returns a list with five vectors containing between one and five random deviates:

map(1:5, runif)

#If the vector 1:5 is a list, the function does not iterate across five values but rather takes one value, the length five vector, as its sole input. Since the input is not a single value, the map function uses the length of the vector as its input:
  
map(list(1:5), runif)
    
# And as a result, the above function is equivalent to: 

map(list(5), runif)

map(5, runif)

map(list(5:9), runif)
```

### Problem 4

What does map(-2:2, rnorm, n = 5) do? Why? What does map_dbl(-2:2, rnorm, n = 5) do? Why?

``` {r 21.5.4}
map(-2:2, rnorm, n = 5)

# The function above takes a length five vector from -2:2 and returns a list of five vectors of five observations along a normal distribution given mean values of -2, -1, 0, 1, and 2, respectively. It takes each value from -2:2 and returns a vector of five observations, looping over the entire vector and storing the vectors in a single list. 

#map_dbl(-2:2, rnorm, n = 5)

# The above function throws the error "Result 1 is not a length 1 atomic vector," because it is unable to store the output of the function as a single value in a vector. Map_dbl expects each iterative output to be a single value, but each iteration of this function is a vector of five observations. In order to return the function's output as a single vector, the map() function would need to be wrapped in unlist(): 

unlist(map(-2:2, rnorm, n = 5))
```

### Problem 5

Rewrite map(x, function(df) lm(mpg ~ wt, data = df)) to eliminate the anonymous function.

``` {r 21.5.5}
# map(x, ~lm(mpg ~ wt, data = .))
```

## 21.9 Other patterns of for loops

### Problem 1

Implement your own version of every() using a for loop. Compare it with purrr::every(). What does purrr’s version do that your version doesn’t?

``` {r 21.9.1}
x <- list(1:5, letters, list(10))


bens_every <- function(x, f) {
out <- vector("logical", 1)
for (. in seq_along(x))
  out <- all((f(.)))
out
}

bens_every(iris, is.character)
bens_every(iris, is.vector)

# Unlike dplyr's every(), my function does not support shortcuts for anonymous functions

iris %>% 
  select_if(is.numeric) %>% 
  every(~mean(.) > 3)
```

### Problem 2

Create an enhanced col_sum() that applies a summary function to every numeric column in a data frame.

``` {r 21.9.2}
col_summary <- function(df, fun) {
  df <- keep(df, is.numeric)
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[[i]] <- fun(df[[i]])
  }
  out
}

col_summary(iris, mean)
```

### Problem 3

A possible base R equivalent of col_sum() is:

``` {}
col_sum3 <- function(df, f) {
  is_num <- sapply(df, is.numeric)
  df_num <- df[, is_num]

  sapply(df_num, f)
}
```

But it has a number of bugs as illustrated with the following inputs:

``` {}
df <- tibble(
  x = 1:3, 
  y = 3:1,
  z = c("a", "b", "c")
)
# OK
boom col_sum3(df, mean)
# Has problems: don't always return numeric vector
col_sum3(df[1:2], mean)
col_sum3(df[1], mean)
#col_sum3(df[0], mean)
```

What causes the bugs?

``` {r 21.9.3}
#The bug stems from sapply, which returns a list rather than a vector when given a dataframe with no variables. Here is a more consistent approach in base R: 

col_sum3 <- function(df, f) {
  is_num <- vapply(df, is.numeric, logical(1))
  df_num <- df[, is_num]

  sapply(df_num, f)
}

df <- tibble(
  x = 1:3, 
  y = 3:1,
  z = c("a", "b", "c")
)

col_sum3(df[0], mean)
col_sum3(df[1:2], mean)
col_sum3(df[1], mean)
```



