---
output:
  pdf_document: default
  html_document: default
---
# Chapter 21 - Iteration {-}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r library, message=FALSE}
library(tidyverse)
library(stringr)
```

## 21.2 For loops {-}

### Problem 1 {-}

Write a for loop to compute the mean of every column in mtcars.

``` {r 21.2.1a}
x <- vector("double", ncol(mtcars))
for (i in seq_along(mtcars)) {
  x[[i]] <- mean(mtcars[[i]])
}
```

Write a for loop to determine the type of each column in nycflights13::flights.
(NOTE: Unsure how to deal with class "unknown" in column 19)

``` {r 21.2.1b}
x <- vector("character", ncol(nycflights13::flights))
for (i in seq_along(nycflights13::flights)) {
  x[[i]] <- str_c(class(nycflights13::flights[[i]]), collapse = ", ")
}
```

Write a for loop to compute the number of unique values in each column of iris.

``` {r 21.2.1c}
x <- vector("integer", ncol(iris))
for (i in seq_along(iris)) {
  x[[i]] <- length(unique(iris[[i]]))
}
```

Write a for loop to generate 10 random normals for each of mean = -10, 0, 10, 100. 
``` {r 21.2.1d}
means <- c(rep(-10, 10), rep(0, 10), rep(10, 10), rep(100, 10))
x <- vector("double", 40)
for (i in seq_along(means)) {
  x[[i]] <- rnorm(1, means[[i]])
}
```

### Problem 2 {-}

Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:

``` {r 21.2.2a}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}

str_c(letters[1:26], collapse = "")
```

``` {r 21.2.2b}
x <- sample(100)
sd <- 0
for (i in seq_along(x)) {
  sd <- sd + (x[i] - mean(x)) ^ 2
}
sd <- sqrt(sd / (length(x) - 1))

sd(x)
```

``` {r 21.2.2c}
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}

cumsum(x)
```

### Problem 3 {-}

Write a for loop that prints() the lyrics to the children’s song “Alice the camel”.

``` {r 21.2.3a}
humps <- c("five", "four", "three", "two", "one", "no")

y <- vector("character", 6)
for (x in seq_along(humps)) {
  x <- str_replace_all(str_c(c(rep("Alice the camel has x humps; ", 3), "So go, Alice, go. "), collapse = ""), "x", humps)
}
print(str_c(c(x, " Now Alice has no humps!"), collapse = ""))
```

Convert the nursery rhyme “ten in the bed” to a function. Generalise it to any number of people in any sleeping structure.

``` {r 21.2.3b}
roll_bed <- function(n, structure) {
  
n <- n:2 
y <- vector("numeric", length(n))
for (x in seq_along(n)) {
  y <- str_c("There were ",n," in the ",structure," and the little one said, 'Roll over! Roll over!' So they all rolled over and  none fell out.")
}
y <- c(y, str_c("There was 1 in the ", structure, " and the little one said, 'Alone at last!'"))
 print(y)
}
roll_bed(10, "bed")
```

Convert the song “99 bottles of beer on the wall” to a function. Generalise to any number of any vessel containing any liquid on any surface.

``` {r 21.2.3c}
beer_wall <- function(n, liquid) {

n <- n:1
y <- vector("numeric", length(n))
for (x in seq_along(n)) {
  y <- str_replace_all(str_c(n, " bottles of beer on the wall, ", n, " bottles of beer. Take one down and pass it around, ", (n-1)," bottles of beer on the wall."), "beer", liquid)
}
n <- n[1]
y <- c(y, str_replace_all(str_c("No more bottles of beer on the wall, no more bottles of beer. Go to the store and buy some more, ", n, " bottles of beer on the wall."), "beer", liquid))
 print(y)
}
beer_wall(10, "beer")
```

