# Chapter 21 - Iteration {-}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r library, message=FALSE}
library(tidyverse)
library(stringr)
library(purrr)
```

## 21.2 For loops {-}

### Problem 1 {-}

Write a for loop to compute the mean of every column in mtcars.

```{r 21-2-1a}
means <- vector(mode = "double", length = ncol(mtcars))

for (i in seq_along(mtcars)) {
  means[[i]] <- mean(mtcars[[i]])
}
```

Write a for loop to determine the type of each column in nycflights13::flights.

```{r 21-2-1b}
types <- vector(mode = "character", length = ncol(nycflights13::flights))

for (i in seq_along(nycflights13::flights)) {
  types[[i]] <- typeof(nycflights13::flights[[i]])
}
```

Write a for loop to compute the number of unique values in each column of iris.

```{r 21-2-1c}
uniques <- vector(mode = "integer", length = ncol(iris))
for (i in seq_along(iris)) {
  uniques[[i]] <- length(unique(iris[[i]]))
}
```

Write a for loop to generate 10 random normals for each of μ=−10, 0, 10, and 100.
```{r 21-2-1d}
mu <- c(-10, 0, 10, 100)

means <- vector(mode = "list", length = length(mu))

for (i in seq_along(mu)) {
  means[[i]] <- rnorm(10, mu[[i]])
}
```

### Problem 2 {-}

Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:

```
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}
```

```{r 21-1-2a}
str_c(letters, collapse = "")
```

```
x <- sample(100)
sd <- 0
for (i in seq_along(x)) {
  sd <- sd + (x[i] - mean(x)) ^ 2
}
sd <- sqrt(sd / (length(x) - 1))
```

```{r 21-1-2b}
x <- sample(100)
sd(x)
```

```
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
```

```{r 21-1-2c}
x <- runif(100)
cumsum(x)
```

### Problem 3 {-}

Write a for loop that `prints()` the lyrics to the children’s song “Alice the camel”.

```{r 21-2-3a}
humps <- c("five", "four", "three", "two", "one", "no")
for (i in seq_along(humps)) {
  for (j in 1:3) {
    print(str_glue("Alice the camel has {humps[i]} humps."))
  }
  if (i < 6) {
    # str_glue is unecessary. it drops [1]
    print(str_glue("So go, Alice, go."))
  } else {
    print(str_glue("Now Alice is a horse"))
  }
}
```

Convert the nursery rhyme “ten in the bed” to a function. Generalise it to any number of people in any sleeping structure.

```{r 21-2-3b}
n_in_the_bed <- function(occupants = 10L) {
  
  for (i in seq(from = occupants, to = 1)) {
    if (i > 1) {
      print(str_glue(
        "There were {i} in a bed
        And the little one said
        'Roll over, roll over'
        So they all rolled over
        And one fell out
        
        "
      ))
    } else if (i == 1) {
      print(str_glue("There was one in a bed
      And the little one said
      'Good night!'"))
    }
  }
}

n_in_the_bed()
```

Convert the song “99 bottles of beer on the wall” to a function. Generalise to any number of any vessel containing any liquid on any surface.

```{r 21-2-3c}
bottles_of_beer <- function(bottles = 99L) {
  for (i in seq(from = bottles, to = 0)) {
    
    bottles_left <- ifelse(i == 0, "no more", i) 
    
    print(str_glue(
      "{i} bottle of beer on the wall, {i} bottle of beer.
      Take one down and pass it around, {bottles_left} bottles of beer on the wall."
      )
    )
  }
}

bottles_of_beer()
```


### Problem 4 {-}

It’s common to see for loops that don’t preallocate the output and instead increase the length of a vector at each step:

```
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output
```

How does this affect performance? Design and execute an experiment.

```{r 21-2-4a}
library(microbenchmark)

grow_vector <- function() {
  
  # create a character vector
  x <- rep("abc", 1000)
  
  # create an empty vector of length 0 for the output
  output <- vector("integer", 0)
  
  # find the length of each observation in x and grow and concatenate it to 
  # output
  for (i in seq_along(x)) {
    output <- c(output, lengths(x[[i]]))
  }
  output
}

microbenchmark(grow_vector())
```

This first chunk grows a vector from length zero to length 1,000. 

```{r 21-2-4b}
preallocate_vector <- function() {
  
  # create a character vector  
  x <- rep("abc", 1000)
  
  # create an empty vector of length length(x) for the output
  output <- vector("integer", length(x))
  
  # find the length of each observation in x and assignment it the corresponding
  # location in the output vector
  for (i in seq_along(x)) {
    output[i] <- lengths(x[[i]])
  }
  output
}

microbenchmark(preallocate_vector())
```

This second chunk preallocates the space in the vector and then populates it. It is an order of magnitude faster. 

## 21.3 - For loop variations {-}

### Problem 1 {-}

Imagine you have a directory full of CSV files that you want to read in. You have their paths in a vector, `files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)`, and now want to read each one with `read_csv()`. Write the for loop that will load them into a single data frame.

```{r 21-3-1, eval=FALSE}
library(tidyverse)

files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)
output <- vector("list", length(files))
for (i in seq_along(files)) {
  output[[i]] <- read_csv(files[[i]])
}
output <- flatten_df(output)
```

### Problem 2 {-}

What happens if you use `for (nm in names(x))` and `x` has no names? 

``` {r 21-3-2a}
x <- c(1, 2, 3)

for (nm in names(x)) {
  print(names(x[nm]))
}
```

What if only some of the elements are named? 

It does not matter. It still iterates along the entire vector. 
 
```{r 21-3-2b}

x <- c("a" = 1, 2, "c" = 3)

for (nm in names(x)) {
  print(names(x[nm]))
}

```

What if the names are not unique?

It works.

``` {r 21-3-2c}
x <- c("a" = 1, "a" = 2, "c" = 3)

for (nm in names(x)) {
  print(names(x[nm]))
}
```

### Problem 3 {-}

Write a function that prints the mean of each numeric column in a data frame, along with its name. For example, `show_mean(iris)` would print:

```
show_mean(iris)
#> Sepal.Length: 5.84
#> Sepal.Width:  3.06
#> Petal.Length: 3.76
#> Petal.Width:  1.20
```

(Extra challenge: what function did I use to make sure that the numbers lined up nicely, even though the variable names had different lengths?)

```{r 21-3-3}

print_means <- function(data = iris, digits = 2) {

  max_str_length <- max(str_length(names(data)))
  
    for (i in seq_along(data)) {
    
    if (is.numeric(data[[i]])) {
      cat("#> ", 
          str_pad(str_c(names(data)[[i]], ": "), width = max_str_length + 2, side = "right"), 
          format(mean(data[[i]]), digits = digits, nsmall = 2),
          "\n",
          sep = "")
    }
  }
}

print_means()
```

### Problem 4 {-}

What does this code do? How does it work?

```
>>>>>>> aaron
trans <- list( 
  disp = function(x) x * 0.0163871,
  am = function(x) {
    factor(x, labels = c("auto", "manual"))
  }
)
for (var in names(trans)) {
  mtcars[[var]] <- trans[[var]](mtcars[[var]])
}
```

The list trans contains two functions: `disp()` and `am()`. `disp()` takes its input and multiplies by `0.0163871` and `am()` takes its input and creates a factor variable with the value "auto" or "manual." The for loop loops over the two functions in the list, uses the variables "disp" and "am" in the dataset mtcars as inputs, and outputs a new mtcars dataset.

## 21.4 For loops vs. functionals {-}

### Problem 1 {-}

Read the documentation for apply(). In the 2d case, what two for loops does it generalise?

Read the documentation for `apply()`. In the 2d case, what two for loops does it generalise?

It generalizes a for loop along a vector inside of a for loop along a series of columns in a data frame or matrix. 

### Problem 2 {-}

Adapt `col_summary()` so that it only applies to numeric columns You might want to start with an `is_numeric()` function that returns a logical vector that has a TRUE corresponding to each numeric column.


```{r 21-4-2}
col_summary <- function(df, fun) {
  
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    if (is_numeric(df[[i]])) {
      out[i] <- fun(df[[i]])
    }
  }
  out
}

col_summary(mtcars, median)
```

## 21.5 The map functions {-}

<<<<<<< HEAD
### Problem 1

Write code that uses one of the map functions to compute the mean of every column in mtcars.

``` {r 21.5.1a}
mtcars %>% 
  map_dbl(mean)
```

Write code that uses one of the map functions to determine the type of each column in nycflights13::flights

``` {r 21.5.1b}
nycflights13::flights %>% 
  map_chr(typeof)
```





Write code that uses one of the map functions to compute the number of unique values in each column of iris

``` {r 21.5.1c}
iris %>% 
  map_int(~n_distinct(.))
```

Write code that uses one of the map functions to generate 10 random normals for each of μ = -10, 0, 10, and 100. 

``` {r 21.5.1d}
c(-10, 0, 10, 100) %>%
  map(rnorm, n = 10)
```

### Problem 2

How can you create a single vector that for each column in a data frame indicates whether or not it’s a factor?

``` {r 21.5.2}
iris %>%
  map_lgl(is.factor)
```

### Problem 3

What happens when you use the map functions on vectors that aren’t lists? What does `map(1:5, runif)` do? Why?

The function iterates across each value, `1:5`, in the vector and returns a list with five vectors containing between one and five random deviates:

``` {r 21.5.3a}
map(1:5, runif)
```

If the vector 1:5 is a list, the function does not iterate across five values but rather takes one value, the length five vector, as its sole input. Since the input is not a single value, the map function uses the length of the vector as its input:
  
```{r 21-5-3b}
map(list(1:5), runif)
```    

And as a result, the above function is equivalent to: 

```
map(list(5), runif)

map(5, runif)

map(list(5:9), runif)
```

### Problem 4

What does `map(-2:2, rnorm, n = 5)` do? Why? What does `map_dbl(-2:2, rnorm, n = 5)` do? Why?

`map(-2:2, rnorm, n = 5)` returns a list with five vectors of length five with values draw from the standard normal distribution with means ranging from -2 to 2. 

`map_dbl(-2:2, rnorm, n = 5)` returns the error `Error: Result 1 is not a length 1 atomic vector` because vectors can't be nested inside of vectors. 

The first works because `map()` returns a list and the second doesn't work because `map_dbl()` returns a vector. 

### Problem 5

Rewrite `map(x, function(df) lm(mpg ~ wt, data = df))` to eliminate the anonymous function.

```
map(x, ~lm(mpg ~ wt, data = .))
```

## 21.9 Other patterns of for loops {-}

### Problem 1

Implement your own version of `every()` using a for loop. Compare it with `purrr::every()`. What does purrr’s version do that your version doesn’t?

``` {r 21.9.1}
x <- list(1:5, letters, list(10))

my_every <- function(x, f) {
  out <- vector("logical", 1)
  for (i in seq_along(x)) {
    out[[i]] <- f(x[[i]])
  }
  !FALSE %in% out
}

x %>% 
  my_every(is.numeric)
```

`purrr:every()` is clever. It doesn't create a vector and store output. Instead, it tests every object and then returns a FALSE when the first vector fails the test. This is more efficient than my code. 

Unlike dplyr's every(), my function does not support shortcuts for anonymous functions

```
iris %>% 
  select_if(is.numeric) %>% 
  every(~mean(.) > 3)
```

### Problem 2

Create an enhanced `col_sum()` that applies a summary function to every numeric column in a data frame.

```{r 21-9-2}
col_summary <- function(df, fun) {
  df_num <- keep(df, is.numeric)
  
  map_dbl(df_num, fun)
}

col_summary(iris, mean)
```

### Problem 3 {-}

A possible base R equivalent of `col_sum()` is:

```
col_sum3 <- function(df, f) {
  is_num <- sapply(df, is.numeric)
  df_num <- df[, is_num]

  sapply(df_num, f)
}
```

But it has a number of bugs as illustrated with the following inputs:


```
df <- tibble(
  x = 1:3, 
  y = 3:1,
  z = c("a", "b", "c")
)
# OK
col_sum3(df, mean)
# Has problems: don't always return numeric vector
col_sum3(df[1:2], mean)
col_sum3(df[1], mean)
col_sum3(df[0], mean)
```
What causes the bugs?

The bug stems from sapply, which returns a list rather than a vector when given a dataframe with no variables. Here is a more consistent approach in base R: 

``` {r 21.9.3}
df <- tibble(
  x = 1:3, 
  y = 3:1,
  z = c("a", "b", "c")
)

col_sum3 <- function(df, f) {
  is_num <- vapply(df, is.numeric, logical(1))
  df_num <- df[, is_num]

  sapply(df_num, f)
}

col_sum3(df, mean)
```