---
title: "Chapter 3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)
```

## 3.2.4

### Problem 1

Run `ggplot(data = mpg)`. What do you see?

```{r 3-2-4-1}
ggplot(data = mpg)
```

We have a blank plot, since we have only constructed the initial plot object.

### Problem 2

How many rows are in `mpg`? How many columns?

```{r 3-2-4-2}
nrow(mpg)
ncol(mpg)
```

There are `r nrow(mpg)` rows and `r ncol(mpg)` columns in the `mpg` dataset.

### Problem 3

What does the `drv` variable describe? Read the help for `?mpg` to find out.

```{r 3-2-4-3}
?mpg
```

The variable `drv` describes the drive of the vehicle: f = front-wheel drive, r = rear wheel drive, 4 = 4wd.

### Problem 4

Make a scatterplot of `hwy` vs `cyl`

```{r 3-2-4-4}
ggplot(mpg, aes(cyl, hwy)) +
  geom_point()
```

### Problem 5

What happens if you make a scatterplot of `class` vs `drv`? Why is the plot not useful?

```{r}
ggplot(mpg, aes(class, drv)) +
  geom_point()
```

Since `class` and `drv` are categorical variables, we 