---
title: "Chapter 5 - Data Transformation"
output: html_document
---

Load the libraries needed for these exercises.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE}
library(tidyverse)
library(nycflights13)
```


## 5.2 - Filter Rows with `filter()`

### Problem 1

Find all flights that:

* Had an arrival delay of two or more hours

```{r 5-2-1a}
flights %>%
  filter(arr_delay >= 120) %>%
  count()
```

* Flew to Houston (IAH or HOU)

```{r 5-2-1-b}
flights %>%
  filter(dest %in% c('IAH', 'HOU')) %>%
  count()
```

* Were operated by United, American, or Delta

```{r 5-2-1c}
flights %>%
  filter(carrier %in% c('UA', 'AA', 'DL')) %>%
  count()
```

* Departed in summer (July, August, and September)

```{r 5-2-1d}
flights %>%
  filter(month %in% c(7, 8, 9)) %>%
  count()
```

* Arrived more than two hours late, but didn’t leave late

```{r d-2-1e}
flights %>%
  filter(arr_delay >= 120, dep_delay <= 0) %>%
  count()
```

* Were delayed by at least an hour, but made up over 30 minutes in flight

```{r 5-2-1f}
flights %>%
  filter(dep_delay >= 60, arr_delay <= dep_delay - 30) %>%
  count()
```

* Departed between midnight and 6am (inclusive)

```{r 5-2-1g}
flights %>%
  filter(dep_time >= 0, dep_time <= 600) %>%
  count()
```

### Problem 2

Another useful dplyr filtering helper is `between()`. What does it do? Can you 
use it to simplify the code needed to answer the previous challenges?

`between()` is a shortcut for `x >= left & x <= right`. We can simplify the last 
answer to Problem 1 as:

```{r 5-2-2}
flights %>%
  filter(between(dep_time, 0, 600)) %>%
  count()
```

### Problem 3

How many flights have a missing dep_time? What other variables are missing? 
What might these rows represent?

We use `is.na()` to filter the flights with a missing departure time.

```{r 5-2-3a}
flights %>%
  filter(is.na(dep_time)) %>%
  count()
```

Using `summary()` to see the breakout of the other variables, there appear to 
be flights that were cancelled.

```{r 5-2-3b}
flights %>%
  filter(is.na(dep_time)) %>%
  summary()
```

### Problem 4

Why is NA ^ 0 not missing? Why is NA | TRUE not missing? 
Why is FALSE & NA not missing? Can you figure out the general rule? 
(NA * 0 is a tricky counterexample!)

Working through these examples:
  * Anything to the zero power is 1
  * Anything OR `TRUE` is `TRUE`
  * Anything AND `FALSE` is `FALSE`

These results apply no matter what the LHS side, and so will apply to NA as well.

```{r 5-2-4a}
NA ^ 0
NA | TRUE
NA & FALSE
```

However operations on NA will return NA. NA * 0 is counterintuitive since 
you would think that anything multiplied by 0 would be 0.

```{r 5-2-4b}
NA * 0
NA ^ 2
NA + 1
```

## 5.3 Arrange Rows with `arrange()`

### Problem 1

How could you use arrange() to sort all missing values to the start? 
(Hint: use is.na()).

We can sort missing values using the format:

```{r 5-3-1}
flights %>%
  arrange(desc(is.na(dep_time))) %>%
  head()
```

### Problem 2

Sort flights to find the most delayed flights. Find the flights that left 
earliest.

The most delayed flights (by `arr_delay`) are:

```{r 5-3-2a}
flights %>%
  arrange(desc(arr_delay)) %>%
  head()
```

The flights that left earliest (by `dep_delay`) are:

```{r 5-3-2b}
flights %>%
  arrange(dep_delay) %>%
  head()
```

### Problem 3

Sort flights to find the fastest flights.

We first calculate average speed in MPH as distance / hours in the air, and sort 
on the calculated variable.

```{r 5-3-3}
flights %>%
  mutate(speed = distance / (air_time / 60)) %>%
  arrange(desc(speed)) %>%
  select(speed) %>%
  head()
```

### Problem 4

Which flights travelled the longest? Which travelled the shortest?

The longest flights are:

```{r 5-3-4a}
flights %>%
  arrange(desc(distance)) %>%
  head()
```

The shortest flights are:

```{r 5-3-4b}
flights %>%
  arrange(distance) %>%
  head()
```

