---
title: "Chapter 3"
output: html_document
---

Load the libraries needed for these exercises.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE}
library(tidyverse)
```

## 3.2

### Problem 1

Run `ggplot(data = mpg)`. What do you see?

We have a blank plot, since we have only constructed the initial plot object 
without any aesthetics.

```{r 3-2-1}
ggplot(data = mpg)
```

### Problem 2

How many rows are in `mpg`? How many columns?

There are `r nrow(mpg)` rows and `r ncol(mpg)` columns in the `mpg` data set.

```{r 3-2-2}
nrow(mpg)
ncol(mpg)
```

### Problem 3

What does the `drv` variable describe? Read the help for `?mpg` to find out.

The variable `drv` describes the drive of the vehicle: f = front-wheel drive, 
r = rear wheel drive, 4 = 4wd.

```{r 3-2-3}
?mpg
```

### Problem 4

Make a scatter plot of `hwy` vs `cyl`.

```{r 3-2-4}
ggplot(mpg, aes(cyl, hwy)) +
  geom_point()
```

### Problem 5

What happens if you make a scatter plot of `class` vs `drv`? Why is the plot 
not useful?

Since `class` and `drv` are categorical variables, we don't see much of a 
meaningful relationship in the scatter plot.

```{r 3-2-5}
ggplot(mpg, aes(class, drv)) +
  geom_point()
```


## 3.3

### Problem 1

What’s gone wrong with this code? Why are the points not blue?

```{r 3-3-1a}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

To set an aesthetic manually, it must go outside of `aes()`.

```{r 3-3-1b}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

### Problem 2

Which variables in mpg are categorical? Which variables are continuous? 
(Hint: type ?mpg to read the documentation for the data set). How can you see 
this information when you run mpg?

We can use the `summary` function to see the mode of each variable.

```{r 3-3-2a}
summary(mpg)
```

Or since the data frame is a tibble, just running `mpg` will show you the 
various variable types.

```{r 3-3-2b}
head(mpg)
```


### Problem 3

Map a continuous variable to color, size, and shape. How do these aesthetics 
behave differently for categorical vs. continuous variables?

Continuous variables will use a gradient to scale `color` and `size`, but will 
throw an error when applied to shape.

```{r 3-3-3a}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy, color = displ)) + 
  geom_point()
```

```{r 3-3-3b}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy, size = displ)) + 
  geom_point()
```

```{r 3-3-3c, error=FALSE}
p <- ggplot(data = mpg, mapping = aes(x = cty, y = hwy, shape = displ)) + 
  geom_point()
```


### Problem 4

What happens if you map the same variable to multiple aesthetics?

Mapping `displ` to `color` and `size` results in the following graph. Not 
necessarily helpful, but two ways of displaying the some variation.
```{r 3-3-4}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy, color = displ, size = displ)) + 
  geom_point()
```

### Problem 5

What does the stroke aesthetic do? What shapes does it work with? 
(Hint: use ?geom_point)

The `stroke` aesthetic will modify the width of the border of a shape. From the 
documentation:

```{r 3-3-5}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(shape = 21, colour = "black", fill = "white", size = 5, stroke = 5)
```

### Problem 6

What happens if you map an aesthetic to something other than a variable name, 
like aes(colour = displ < 5)?

In this case the condition we pass to `color` returns a boolean that will map 
to `color`.
```{r 3-3-6}
ggplot(mtcars, aes(wt, mpg, color = disp < 100)) +
  geom_point()
```

